Empirical Bayes estimation of the normal inverse Gaussian prior in simultaneous-equation models

### 2019-05-13 docs/manuscript1.Rnw
changed some nomenclature and added nigig model

	\section{Introduction}
	- Motivation

@@

  The normal inverse Gaussian (NIG) prior model for $\beta_{jd}$ and 
  $\gamma^2_d$ in (\ref{eq:priormodel}) was introduced in 
  \cite{barndorff-nielsen_hyperbolic_1978} and is, since 
  \cite{barndorff-nielsen_normal_1997}, routinely applied in mathematical 
  finance (see, e.g., \cite{kalemanova_normal_2007}). In Figure 
  \ref{fig:dens_igaussian_marginalbeta}, the NIG 
  prior distribution marginalised over $\gamma_d^2$ is depicted together with 
  three other common prior choices,
  Student's $t$ (see Section \ref{sec:inversegammamodel} for more details on 
  this prior), ridge and lasso, all scaled to variance one. From this Figure 
  we see that, depending on the choice
  of hyperparameters, the NIG prior can vary from heavier tails than the lasso, 
  to more Gaussian tails like the ridge, but behaves similarly to the 
  Student's $t$ prior. Our argumentation to model the 
  $\gamma^2_d$ by an inverse Gaussian 
  distribution, as has been suggested in \cite{fabrizi_specification_2016} and 
  \cite{caron_sparse_2008}, is two-fold: (i) the NIG model is more flexible than
  the standard ridge and lasso priors (as seen from Figure 
  \ref{fig:dens_igaussian_marginalbeta}), and (ii) the NIG prior allows to model
  the mean $\theta_d$ as a 
  function of the co-data $\mathbf{c}_d$ more conveniently than the Student's 
  $t$ prior, as explained in Section 
  \ref{sec:empiricalbayes}. 
<<dens_igaussian_marginalbeta, fig.cap="Several choices of marginal $\\beta_j$ prior, scaled to $\\V(\\beta_j)=1$.", out.width="80%", fig.asp=2/3>>=
library(GeneralizedHyperbolic)
library(metRology)
# all scaled such that variance is one
sigma <- 1
theta <- 1/sigma^2
lambda <- c(0.1, 2)
lambda1 <- sqrt(2)
eta <- lambda/theta + 2
col <- c(1:6)
lty <- c(1:6)
labels <- as.expression(c(bquote("NIG, "~lambda==.(lambda[1])~", "~
                                   theta==.(theta)),
                          bquote("NIG, "~lambda==.(lambda[2])~", "~
                                   theta==.(theta)),
                          bquote("Student's t, "~lambda==.(lambda[1])~", "~
                                   eta==.(eta[1])),
                          bquote("Student's t, "~lambda==.(lambda[2])~", "~
                                   eta==.(eta[2])),
                          "ridge", "lasso"))
dprior <- function(x, lambda, theta, sigma) {
  dnig(x, 0, sigma*sqrt(lambda), sqrt(lambda/(theta*sigma)), 0)
}
dlasso <- function(x, lambda1) {
  0.5*lambda1*exp(-lambda1*abs(x))
}

@@

  \section{Simulations}
  \subsection{Comparison of NIG to normal inverse Gamma model}
  \label{sec:inversegammamodel}
  A common default prior for latent variance components is the inverse Gamma
  prior:
  $$
  \gamma_d^2 \sim \Gamma^{-1}(\eta_d/2,\lambda_d/2),
  $$
  with shape $\eta_d/2$ and scale $\lambda_d/2$, both rescaled to match our 
  inverse Gaussian prior formulation. The resulting marginal prior for the model
  parameters
  (conditional on the error variance $\sigma_d^2$) is then a scaled Student's 
  $t$-distribution:
  $$
  \beta_j | \sigma_d^2 \sim \text{Student's } 
  t\(\eta_d,\sigma_d \sqrt{\lambda_d/\eta_d}\),
  $$
  with degrees of freedom $\eta_d$ and scale $\sigma_d \sqrt{\lambda_d/\eta_d}$.
  See SM Section \ref{sm-sec:inversegamma} for more details on this prior. 
  We compare the 
  posteriors of the proposed NIG prior to the scaled Student's $t$-distribution
  prior in a simulated setting. We generate draws from these posteriors using 
  an MCMC sampler implemented in \texttt{rstan}. The aim of this simulation 
  study is 
  to investigate the performance of the NIG prior model, so we set $D=1$ and 
  drop the $d$ indices in the following.

@@

  To assess the performance of the variational approximations we evaluate 
  several measures. Divergence between the MCMC and variational posterior
  is quantified through a multivariate extension of the Cram{\'e}r-von Mises
  criterion \cite[]{baringhaus_new_2004}, where smaller values of the criterion
  indicate more similar distributions. In addition, we calculate the correlation
  and mean squared distance (MSD) between the posterior variances of
  the MCMC posterior and variational posterior. \texttt{rstan} allows the user 
  to maximise the posterior to obtain the maximum \textit{a posteriori} (MAP)
  estimates of the model parameters. We include these MAP estimates together
  with the ridge estimates obtained from the \texttt{glmnet} \texttt{R} package
  \cite[]{friedman_regularization_2010} in the comparison of the posterior mean
  estimates.

  The interest of this simulation lies in the variational Bayes approximation, 
  so we restrict ourselves to the univariate setting (i.e., $D=1$) and drop any
  $d$ indices from notation in the following. We simulated
  $n=100$ features $\x_i$ from a $p=200$ dimensional Gaussian distribution with
  zero mean vector, variances equal to one, and all covariances set to 
  $\rho=0.5$. The outcomes are generated through $y_i = \x_i\tr \bbeta + 
  \epsilon_i$, with the $\epsilon_i$ from the standard normal distribution. 
  Likewise, we generate the model parameters $\beta_j$ from a standard normal
  distribution. For estimation, we set our hyperparameters $\lambda=0.1$ and 
  $\theta=2$. We choose the penalty parameter in \texttt{glmnet} by assuming a 
  known error variance and matching the marginal prior variance of the NIG and
  ridge models, which results in the penalty parameter 
  $1/(n\sigma^2\theta)=1/200$.To mitigate the effect of random variation, we
  repeat the simulation 100 times and report the results in Figures 
  \ref{fig:boxplots_igaussian_res3_vb}-
  \ref{fig:boxplots_igaussian_res3_best}.

@@

@@

  \section{Application}

>>>>

	\section{Introduction}
	- Motivation:
	simultaneous regressions in: network reconstruction, eQTL mapping 
	\cite[]{kpogbezan_incorporating_2019}, drug target discovery 
	\cite[]{noh_inferring_2016}.

	- Introduce data example

@@

  The normal inverse Gaussian (NIG) prior model for $\beta_{jd}$ and 
  $\gamma^2_d$ in (\ref{eq:priormodel}) was introduced in 
  \cite{barndorff-nielsen_hyperbolic_1978} and is, since 
  \cite{barndorff-nielsen_normal_1997}, routinely applied in mathematical 
  finance (see, e.g., \cite{kalemanova_normal_2007}). See SM Section 
  \ref{sm-sec:nigprior} for more details on this prior.
  In Figure \ref{fig:dens_igaussian_marginalbeta}, the NIG 
  prior distribution marginalised over $\gamma_d^2$ is depicted together with 
  three other common prior choices,
  Student's $t$ (see Section \ref{sec:inversegammamodel} for more details on 
  this prior), ridge and lasso, all scaled to variance one. From this Figure 
  we see that, depending on the choice
  of hyperparameters, the NIG prior can vary from heavier tails than the lasso, 
  to more Gaussian tails like the ridge, but behaves similarly to the 
  Student's $t$ prior. Our argumentation to model the 
  $\gamma^2_d$ by an inverse Gaussian 
  distribution, as has been suggested in \cite{fabrizi_specification_2016} and 
  \cite{caron_sparse_2008}, is two-fold: (i) the NIG model is more flexible than
  the standard ridge and lasso priors (as seen from Figure 
  \ref{fig:dens_igaussian_marginalbeta}), and (ii) the NIG prior allows to model
  the mean $\theta_d$ as a 
  function of the co-data $\mathbf{c}_d$ more conveniently than the Student's 
  $t$ prior, as explained in Section 
  \ref{sec:empiricalbayes}. 
<<dens_igaussian_marginalbeta, fig.cap="Several choices of marginal $\\beta_j$ prior, scaled to $\\V(\\beta_j)=1$.", out.width="80%", fig.asp=2/3>>=
library(GeneralizedHyperbolic)
library(metRology)
# all scaled such that variance is one
sigma <- 1
theta <- 1/sigma^2
lambda <- c(0.1, 2)
lambda1 <- sqrt(2)
eta <- lambda/theta + 2
col <- c(1:6)
lty <- c(1:6)
labels <- as.expression(c(bquote("NIG, "~lambda==.(lambda[1])~", "~
                                   theta==.(theta)),
                          bquote("NIG, "~lambda==.(lambda[2])~", "~
                                   theta==.(theta)),
                          bquote("Student's t, "~lambda==.(lambda[1])~", "~
                                   eta==.(eta[1])),
                          bquote("Student's t, "~lambda==.(lambda[2])~", "~
                                   eta==.(eta[2])),
                          "ridge", "lasso"))
dprior <- function(x, lambda, theta, sigma) {
  dnig(x, 0, sigma*sqrt(lambda), sqrt(lambda/(theta*sigma)), 0)
}
dlasso <- function(x, lambda1) {
  0.5*lambda1*exp(-lambda1*abs(x))
}

@@

  \section{General model}
  In addition to the outcome co-data $\mathbf{c}_d$, we may have feature co-data
  $\mathbf{z}_j$ available. Examples of this in omics settings are pathway 
  memberships or previously obtained $p$-values for the features. In
  that case, we extend model (\ref{eq:priormodel}) to:
  \begin{subequations}\label{eq:generalmodel}
    \begin{align}
      y_{id} | \bm{\beta}_d, \sigma_d^2 & \sim \mathcal{N} 
      (\x_i \tr \bm{\beta}_d, \sigma_d^2), \nonumber \\
      \beta_{jd} | \tau_j^2, \gamma_d^2, \sigma_d^2 & \sim \mathcal{N} 
      (0, \sigma_d^2 \gamma_d^2 \tau_j^2), \\
      \gamma_d^2 & \sim \mathcal{IG}\(\theta_d^c, \lambda^c_d\), \\
      \tau_j^2 & \sim \mathcal{IG}\(\theta_j^z, \lambda^z_j\), \\
      \sigma_d^2 & \sim 1/\sigma_d^3, \nonumber
    \end{align}
  \end{subequations}
  where we now have an extra variance component $\tau_j$ that is 
  \textit{a priori} inverse Gaussian distributed with mean dependent on the 
  feature co-data $\mathbf{z}_j$. Equivalently, we may set 
  $\eta_{jd}^2=\tau_j^2 \gamma_d^2/(\theta_j^z \theta_d^c)$ and write 
  (\ref{eq:generalmodel}) as:
  \begin{align*}
    \beta_{jd} | \eta_{jd}^2, \sigma_d^2 & \sim \mathcal{N} 
    \(0, \frac{\lambda_{jd}}{\chi_{jd}}\eta_{jd}^2\sigma_d^2\), \\
    \eta_{jd}^2 & \overset{D}{=}\pi^{-1}\sqrt{\chi_{jd}}\exp(\psi_{jd}) 
    \eta_{jd}^{-3/2} \cdot K_{0} \(\sqrt{\psi_{jd}^2 + \chi_{jd} (\eta_{jd}^2 
    + \eta_{jd}^{-2} - 2)} \),
  \end{align*}
  where $\lambda_{jd}=\lambda_j^z\lambda_d^c$, 
  $\chi_{jd}=\lambda_j^z\lambda_d^c/(\theta_j^z \theta_d^c)$, and
  $\psi_{jd}=\lambda_j^z/\theta_j^z + \lambda_d^c/ \theta_d^c$. A third way
  of writing the model is:
  \begin{subequations}\label{eq:generalprior}
    \begin{align}
      \beta_{jd} | \eta_{jd}^2, \sigma_d^2 & \sim \mathcal{N} 
      \(0, \eta_{jd}^2 \sigma_d^2\), \\
      \eta_{jd}^2 | \xi_{d}^2 & \sim \mathcal{IG} 
      (\theta_j^z \xi_d^2, \lambda_j^z \xi_d^2 ), \\
      \xi_d^2 & \sim \mathcal{GIG} (1, \lambda_d^c/(\theta_d^c)^2, \lambda_d^c).
    \end{align}
  \end{subequations}
  From (\ref{eq:generalprior}) we find the marginal variance and kurtosis for 
  $\beta_{jd}$ (conditional on $\sigma_d^2$):
  \begin{align}
    \V(\beta_{jd} | \sigma_d^2) & =\sigma_d^2 \theta_j^z \theta_d^c 
    \[\frac{K_0(\lambda_d^c/\theta_d^c)}{K_1(\lambda_d^c/\theta_d^c)} + 
    \frac{2\theta_d^c}{\lambda_d^c}\],\nonumber \\
    \mathcal{K}(\beta_{jd} | \sigma_d^2) & = 3 \(\frac{\theta_j^z}{\lambda_j^z}
    + 1\) \theta_d^c \[r_d - r_d^{-1} + 8\(\frac{\theta_d^c}{\lambda_d^c}\)^2
    K_0 (\lambda_d^c/\theta_d^c) + \frac{4 \theta_d^c}{\lambda_d^c}r_d 
    K_0 (\lambda_d^c/\theta_d^c) + 1/\theta_d^c\], \label{eq:kurtnigig}
  \end{align}
  with $r_d = K_0(\lambda_d^c/\theta_d^c)/K_1(\lambda_d^c/\theta_d^c)$. If 
  we fix the prior expectation $\E(\gamma_d^2)=\theta_d^c$ and let the prior 
  variance 
  $\V(\gamma_d^2)$ go to zero, we have $\V(\beta_{jd} | \sigma_d^2) \rightarrow
  \sigma_d^2 \theta_j^z \theta_d^c$ and $\mathcal{K}(\beta_{jd} | \sigma_d^2)
  \rightarrow 3(\theta_j^z/\lambda_j^z + 1)$. From (\ref{eq:kurtnigig}) we see
  that we can keep the variance constant by fixing $\theta_j^z$, $\theta_d^c$,
  and $\lambda_d^c$, while we increasing the kurtosis by decrearing 
  $\lambda_j^z$. This makes the model (\ref{eq:generalmodel}), which we 
  refer to by the NIGIG model hereafter, a flexible model.

  The formulation as in (\ref{eq:generalmodel}) results in an additional 
  variational distribution for the $\tau_j$:
  $$
  q(\tau_j^2) \overset{D}{=} \mathcal{GIG}\(-\frac{D+1}{2}, \lambda^z_j 
  (\theta_j^z)^{-2}, \delta_j^z\).
  $$
  The variational parameters are now updated by:
  \begin{align*}
    \bm{\Sigma}_d^{(h+1)} & = (a_d^{(h)})^{-1} 
    \[\X \tr \X + b_d^{(h)} \diag (c_j^{(h)})\]^{-1}, \\
    \bm{\mu}_d^{(h+1)} & = \[\X \tr \X + b_d^{(h)} \diag(c_j^{(h)})\]^{-1} 
    \X \tr \y_d, \\
    (\delta^c_d)^{(h+1)} & = a_d^{(h)} \left\{(\bmu^{(h+1)}_d) \tr 
    \diag (c_j^{(h)}) \bmu^{(h+1)}_d + \trace \[ \diag (c_j^{(h)}) 
    \bSigma_d^{(h+1)}\] \right\} + \lambda^c_d, \\ 
    (\delta^z_j)^{(h+1)} & = \sum_{d=1}^D a_d^{(h)} b_d^{(h)} 
    \[(\bmu^{(h+1)}_d) \tr \bmu^{(h+1)}_d + \trace (\bSigma_d^{(h+1)}) \] + 
    \lambda^z_j, \\ 
    \zeta_d^{(h+1)} & = \frac{1}{2} \bigg[\mathbf{y}_d \tr \mathbf{y}_d -
    2 \mathbf{y}_d \tr \X \bm{\mu}_d^{(h+1)} + 
    \trace ( \X \tr \X \bm{\Sigma}_d^{(h+1)}) + 
    (\bm{\mu}_d^{(h+1)}) \tr \X \tr \X \bm{\mu}_d^{(h+1)} \\
    & \,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\, + 
    b_d^{(h+1)} \trace \[ \diag (c_j^{(h)}) \bm{\Sigma}_d^{(h+1)}\] + 
    b_d^{(h+1)} (\bm{\mu}_d^{(h+1)}) \tr \diag (c_j^{(h)}) 
    \bm{\mu}_d^{(h+1)}\bigg],
  \end{align*}
  until convergence, where
  $$
  c_j^{(h)} = \E_{Q^{(h)}}(\tau_j^{-2}) = \sqrt{\frac{\lambda^z_j}
  {(\theta_j^z)^2 (\delta^z_j)^{(h)}}} 
  \frac{K_{\frac{D - 1}{2}} 
  \( \sqrt{\lambda^z_j (\delta^z_j)^{(h)}}/\theta_j^z \)}
  {K_{\frac{D+1}{2}} \( \sqrt{\lambda^z_j (\delta_j^z)^{(h)}}/\theta_j^z \)} +
  \frac{D+1}{(\delta_j^z)^{(h)}}
  $$
  If we set $\theta_d^c=(\mathbf{c}_d \tr \bm{\alpha}^c)^{-1}$ and
  $\theta_j^z=(\mathbf{z}_j \tr \bm{\alpha}^z)^{-1}$, we have an additional
  empirical Bayes step that consists of an iteratively 
  reweighted least squares regression of responses $(f_j^{(l)})^{-1}$ on
  predictors $\mathbf{z}_j$ with weights $\lambda_j^z f_j^{(l)}$, where 
  \begin{align*}
  f_j^{(l)} & = \E_{Q^{(h)}}(\tau_j) = [c_j^{(l)} - 
  (D + 1)/(\delta_j^z)^{(l)}] \cdot (\delta_j^z)^{(l)} [\mathbf{z}_j \tr 
  (\bm{\alpha}^z)^{(l)} ]^{-2} / (\lambda_j^z)^{(l)}, \\
  \lambda_j^z & = \[ c_j^{(l)} + f_j^{(l)} 
  (\mathbf{z}_j \tr \bm{\alpha}^z)^2 - 
    2\mathbf{z}_j \tr \bm{\alpha}^z \]^{-1}.
  \end{align*}

  \section{Simulations}
  \subsection{Comparison of NIG and Student's \texorpdfstring{$t$}{t} priors}
  \label{sec:inversegammamodel}
  A common default prior for latent variance components is the inverse Gamma
  prior:
  $$
  \gamma_d^2 \sim \Gamma^{-1}(\eta_d/2,\lambda_d/2),
  $$
  with shape $\eta_d/2$ and scale $\lambda_d/2$, both rescaled to match our 
  inverse Gaussian prior formulation. The resulting marginal prior for the model
  parameters
  (conditional on the error variance $\sigma_d^2$) is then a scaled Student's 
  $t$-distribution:
  $$
  \beta_j | \sigma_d^2 \sim \text{Student's } 
  t\(\eta_d,\sigma_d \sqrt{\lambda_d/\eta_d}\),
  $$
  with degrees of freedom $\eta_d$ and scale $\sigma_d \sqrt{\lambda_d/\eta_d}$.
  See SM Section \ref{sm-sec:inversegamma} for more details on this prior. 
  We compare the 
  posteriors of the proposed NIG prior and the scaled Student's $t$-distribution
  prior in a simulated setting.

  We compare the two posterior distributions to the true posterior for a 
  number of measures. Divergence between the distributions
  is quantified through a multivariate extension of the Cram{\'e}r-von Mises
  criterion \cite[]{baringhaus_new_2004}, where smaller values of the criterion
  indicate more similar distributions. We also assess 95\% coverage of the 
  posterior credible intervals. The accuracy of the 
  point estimates of $\beta_j$ in the form of posterior means is assessed by
  the mean squared distance (MSD) and correlation between the point estimates 
  and the true $\beta_j$.

  The aim of this simulation study is to investigate the performance of the NIG 
  prior model, so we set $D=1$ and drop the $d$ indices in the following. 
  We studied two settings: one dense and one sparse. In the dense setting, 
  we simulated $n=100$ features $\x_i$ from a $p=200$ dimensional Gaussian 
  distribution with zero mean vector, variances equal to one, and all 
  covariances set to $\rho=0.5$. The outcomes were generated through $y_i = 
  \x_i\tr \bbeta + \epsilon_i$, with the $\epsilon_i$ from the standard normal 
  distribution. 
  Likewise, we generated the model parameters $\beta_j$ from a standard normal
  distribution. For estimation, we set our hyperparameters $\lambda=0.5$, 
  $\theta=1$, and $\eta=2.5$. such that the marginal $\beta_j$ variances
  of the NIG and Student's $t$ priors match. For both priors, we generated 
  1000 draws (after 1000 burnin) from the posteriors using an MCMC sampler 
  implemented in \texttt{rstan}.

  In the sparse setting, the observed data generating process is the same. The
  model parameters are are simulated from a spike-and-slab prior with a 
  Laplace slab:
  \begin{align*}
    w & \sim \text{Beta}(\nu, \kappa), \\
    z_j | w & \sim \text{Binom}(w), \\
    \beta_j | z_j & \sim z_j\text{Laplace}
    \(\sqrt{\frac{\nu + \kappa}{2\nu}} \sigma \gamma\) + (1 - z_j)\delta_0,
  \end{align*}
  where $\delta_0$ denotes the Dirac mass at zero. The Laplace scale is chosen
  such that the we have marginal prior variance $\V(\beta_j)=1$. The sparsity 
  level is determined by the proportion of zero $\beta_j$, 
  $s=1 - \nu/(\nu + \kappa)$. Note that the resulting marginal prior for
  $\beta_j$ is:
  $$
  \beta_j \sim (s - 1)\text{Laplace}\(\sqrt{\frac{\nu + \kappa}{2\nu}} 
  \sigma \gamma\) + s\delta_0.
  $$
  We use two settings, $s=0.5$ and $s=0.9$. The 
  hyperparameters for estimation are set the same as in the dense setting.
  To mitigate the effect of random variation, we
  repeated the simulations 100 times and report the results in Figures 
  \ref{fig:boxplots_igaussian_res4_post} and [! ref to figure here]
<<boxplots_igaussian_res4_post, fig.cap="Comparison of the NIG and Student's t priors through (a) Cramér-von Mises criterion between the model and true posteriors, (b) correlation and (c) MSD between the posterior means and true parameters, and (d) 95\\% credible interval coverage of the true values.", out.width="100%", fig.asp=2/5>>=
@

  [! describe results here]

@@

  To assess the performance of the variational approximations we evaluate 
  several measures. We measure divergence between the MCMC and variational 
  posterior with the Cram{\'e}r-von Mises criterion. We compute the 
  correlation and MSD between the MCMC and variational posterior variances. 
  \texttt{rstan} allows the user 
  to maximise the posterior to obtain the maximum \textit{a posteriori} (MAP)
  estimates of the model parameters. We include these MAP estimates together
  with the ridge estimates obtained from the \texttt{glmnet} \texttt{R} package
  \cite[]{friedman_regularization_2010} in the comparison of the posterior mean
  estimates.

  The interest of this simulation lies in the variational Bayes approximation, 
  so we restrict ourselves to the univariate setting (i.e., $D=1$). The
  simulation setup is the same as in Section \ref{sec:inversegammamodel}.
  The hyperparameters are set as follows: $\lambda=0.1$ and 
  $\theta=2$. We choose the penalty parameter in \texttt{glmnet} by assuming a 
  known error variance and matching the marginal prior variance of the NIG and
  ridge models, which results in the penalty parameter 
  $1/(n\sigma^2\theta)=1/200$. We
  repeated the simulation 100 times and report the results in Figures 
  \ref{fig:boxplots_igaussian_res3_vb}-
  \ref{fig:boxplots_igaussian_res3_best}.

@@

  [! describe results here]

@@

  \section{Application}
  - Possible: network reconstruction, medical surveys, drug response prediction
